Draft: It's Not So Bad---The minimal effect of election fraud on protest intensity in non-democracies

### 2022-07-20 Output/manipulation_protest_draft_v2.Rmd
Finished rough draft of v3 of paper (DV: quantity of protest)

---
title: "Draft: Judicial independence and post-election protest"
author: "Cole J. Harvey (Oklahoma State University)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    bookdown::pdf_document2: default
linkcolor: blue
bibliography: C:/Users/colej/Documents/Research projects/bib_ch.bib
#biblio-style: "apalike"
fontsize: 12pt
link-citations: yes
toc: false
csl: C:/Users/colej/Documents/Research projects/ji_manipulation_protest/american-journal-of-political-science.csl
header-includes: \usepackage{booktabs}
abstract: "Multiparty elections and courts with a degree of independence are common features of non-democratic politics. Previous research posits a connection between these two institutions, arguing that judicial independence increases the risk of post-election protest, thus reducing electoral manipulation. This paper argues that the relationship is not straightforward, identifying four distinct (and competing) channels through which courts may affect protest risk. Using cross-national data from 1990-2012, the paper shows that independent courts only increase protest risk under limited circumstances. The results show that judicial independence increases protest risk where economic grievances are high, while reducing protest risk where grievances are low. As a result, creates judicial independence an underappreciated risk of democratic backsliding in well-off societies, while improving the odds of democratization elsewhere."
---

@@

Using cross-national data on 467 elections from 1990 to 2012, adapted from the recently-released Electoral Contention and Violence (ECAV) dataset [@Daxecker2019], this paper finds that more fraudulent elections are associated with a higher risk any post-election protest event occurring. This is in keeping with the formal-modeling literature, which predicts that overreach in the form of election fraud indicates government weakness, making opposition groups confident of winning a conflict in the streets. However, there is no statistically significant association between the level of election fraud and the duration of protest. This suggests a more complex relationship between election fraud and protest. One the one hand, the result is consistent with the prediction of some formal models that fraud is a noisy signal---in some cases, opposition groups may misperceive a strong government as weak. On the other hand, since fraud is not associated with the overall duration of a protest wave, the risk to governments has likely been overemphasized. That is, even if protests do break out in response to large-scale fraud, the government has many tools it may deploy to curtail them. To the extent that longer protest waves are more damaging to the government than short ones, protest risk does not appear to be a substantial constraint on the severity of election fraud.

@@

Sustained protests may also contribute to elites' decisions to stand by the incumbent regime, or to side with "the people" in their own pursuit of power [@hale_patronal_2014;@svolik_politics_2012;@robertson_protest_2010].

The literature thus expects that manipulated elections increase the risk of protest by 1) revealing regime weakness, especially when the margin of victory is low, and 2) exacerbating existing grievances and upsetting citizens' sense of procedural fairness. However, as discussed below, electoral protest is relatively rare, and even in cases where it does occur, most efforts to overturn fraudulent election results through protest have failed [@kalandadze2009electoral]. The apparent failure of protest risk to deter election fraud underlines our limited understanding of non-democratic ruling parties' decision-making. Does the extent of fraud, or the severity of underlying social grievances, make long-lasting protest more likely (and thus restrain fraud in anticipation)?

@@

Research on the duration of protest waves is limited [@Davenport2019], but the foregoing discussion suggests several baseline hypotheses to be tested in the the occurrence of post-election protest as a binary variable. These include, first, the prediction that more severe election fraud is a signal of deeper regime weakness (i.e. unpopularity), which makes protest more likely. Second, there is the alternative view that election manipulation is most likely to spark a protest movement when margins of victory are narrow [@Simpser2011]. Finally, the grievances approach suggests that fraud will most likely lead to protest when social grievances are high [@Brancati2016;@tucker2007].

@@

However, these predictions may not apply to the duration of a protest wave. Once protests begin, the extent of electoral fraud is unlikely to affect their duration. The initial emergence of protest triggers a response by the government and ruling party that can take the form of repression, concessions, or toleration [@Tilly1977;@Lipsky1968;@Hummel2019;@Rasler1996;@Bishara2015;@Frantz2014]. (**See others**.) Given that post-electoral protests are a direct challenge to the regime, toleration is unlikely in these cases. That election dates are known in advance creates opportunities for opposition mobilization; anticipating this, regimes can more easily engage in pre-emptive repression [@Truex2019;@Sullivan2016;@Bhasin2013]. Because the emergence of protest indicates that the regime's pre-emptive efforts have failed, it is not clear a priori how the regime should handle protesters it failed to deter [@Ritter2016]. In many cases, the government doubles down on repression, calling upon the security services to disperse protests, arrest leaders, and perhaps employ harsher methods including disappearance, torture, and execution [@Escriba_Folch2013;@Davenport2007]. Though such crackdowns risk provoking a backlash, they are aimed at demobilizing the opposition by increasing the costs of participation in overt dissent [@DeVogel2022]. In other cases, the government may attempt to accommodate the opposition, by making concessions on policy or institutions [@Moore2000;@Hafner-Burton2016;@Piven1979]. Alternatively, concessions may be aimed at the more moderate public, in order to keep them on the sidelines of the conflict [@Hummel2019].

@@

In this struggle between the potential protest coalition and the incumbent government, protesters and their possible allies behave as "passionate economists" [@VanZomeren2012]---they are motivated both by emotional responses to grievance, and by a more rational calculation of individual-level risk and group-level efficacy. What is the likelihood that protest action can help resolve the grievance, and does the benefit of participating outweigh the risk of personal consequences to the individual for participating?  A fraudulent election may heighten citizens' grievances and generate uncertainty around the underlying popularity of the regime, creating a moment of elevated protest risk. Crucially, this evaluation is dynamic, and can be updated over time in response to new information and new events [@VanZomeren2012]. If protest movements have sufficient numbers and organizational resources to put the regime on the back foot, this can lead to more durable protest movements by increasing individuals' sense of group efficacy and decreasing their sense of risk [@Chenoweth_Stephan2011]. On the other hand, if the regime is well-resourced, capable, and unified, its actions to end a protest movement (including repression, concessions, and co-optation) can lead to a negative re-evaluation of the movement's odds of success. This does not, of course, eliminate citizens' sense of grievance. Instead, it leads them to abandon 'approach-oriented' forms of coping, to avoidance-oriented forms: demobilization and disengagement [@van_zomeren_put_2004;@Young2019]. For example, while ruling-party supporters in Russia disapprove of election fraud [@Reuter2021], they are more likely to withdraw from politics after exposure to electoral protest than to support the opposition [@Tertytchnaya2020].

In summary, though election fraud may be taken as a signal by opposition groups, it is a noisy one: weak and strong governments alike may engage in election manipulation. However, once a protest movement begins, the level of election manipulation is no longer a useful signal of incumbent strength. Instead, protesters and bystanders can update their estimation of the relative strength of the protest coalition and the incumbent government by observing the size of crowds, the number of protests, and the government's reaction. Since the level of manipulation will no longer serve as an important indicator of regime strength once a challenge is taken to the streets, the duration of protests will instead be related to other indicators of opposition and incumbent capacity. As identified in prior research, these factors include the previous repressive history of the government [@Davenport2019], the institutionalization of opposition parties [@trejo_ballot_2014], the margin of victory in the election [@simpser2013], and whether or not the chief executive faces term limits [@hale_patronal_2014].

>H4: Once protests begin, the severity of election fraud will not be correlated with the duration of protest.

@@

ECAV, from which study's dependent variable is drawn, records protests and other contentious actions from six months before to three months after an election. For this study, I include protests only, excluding acts like shootings, bombings, and coups. ECAV data is in event-day format. To capture *protest duration*, I take the number of days between the first electoral protest and the last. The result is a count variable ranging from 0 (no protests) to 90. The large majority of elections experience no post-election protest: in 460 of the 647 elections, no protests are recorded. Figure 1 reports the distribution of protest duration across the dataset, as well as winners' vote-shares.

@@

This distribution is in line with the theoretical proposition that two data-generating processes are at work: the factors that lead to the occurrence of protest may not be the same as those that sustain or intensify it. Consequently, I model the results in two ways. The first approach uses logit models to identify the factors that make *any protest* more likely to occur---this is a binary variable that is coded as 1 if any protest occurred after the election. Since election-year observations are nested within countries here, I use multilevel models with random intercepts by country in. In the second analysis, I use negative binomial regressions to model the duration of protests given that protest has begun. That is, these latter models exclude cases where no protest occurred. For ease of interpretation, these processes are modeled and depicted separately in the main text, while hurdle regression models (which use the full dataset in one stage) are presented in the appendix, with similar results.

@@

Next, all models include three measures of the degree of political contestation in the country. The first of these is an indicator variable for *electoral democracy*; the reference category is electoral autocracy. The second and third are *judicial constraints* and  *legislative constraints* on the executive, V-Dem variables which captures the extent to which the judicial and legislative bodies in the country act independently in practice to limit executive power. Since urban areas can make electoral manipulation more difficult [@ziblatt2009;@larreguy2016parties] while also making protest more likely [@Wallace2013a], I also include data on *urbanization* taken from the United Nations [@DepartmentofEconomicandSocialAffairsPopulationDivision2019].

@@

```{r count-models-interact, warning = FALSE, message = FALSE, include = F}
model.d.base2 <- glm.nb(d.events.num ~ jucon.1lag + legcon.1lag + gdpgro.1lag + pubseccorrup.lag + osorg.1lag + econ.crisis.nelda.1lag + natlpartyorg.lag + v2elirreg.inv + pubseccorrup.lag +    winner.share.upd*v2elirreg.inv  + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + rentierstate + taxstate, data = protest.cases %>% filter(gdpgro.1lag < 1))
#summary(model.d.base2)
p2 <- sjPlot::plot_model(model.d.base2, type = "pred", terms = c("winner.share.upd [all]", "v2elirreg.inv [-.36, 1.85]")) + coord_cartesian(ylim = c(0, 200))

model.n.base2 <- glm.nb(n.events.post ~ jucon.1lag + legcon.1lag + gdpgro.1lag + pubseccorrup.lag + osorg.1lag + econ.crisis.nelda.1lag + natlpartyorg.lag + v2elirreg.inv + pubseccorrup.lag +   winner.share.upd*v2elirreg.inv  + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + rentierstate + taxstate, data = protest.cases %>% filter(gdpgro.1lag < 1))
#summary(model.n.base2)
sjPlot::plot_model(model.n.base2, type = "pred", terms = c("winner.share.upd [all]", "v2elirreg.inv [-1, 1.79]"))

@@

```{r tables-quantity, include=FALSE}
models.quantity <- list("Model 13" = model.n.base, "Model 14" = model.n.base2, "Model 15" = model.n.griev, "Model 16" = model.n.griev2, "Model 17" = model.n.griev3)
#modelsummary(tidytest)
modelsummary(models.quantity, stars = T)
```

@@

```{r tables-duration, results='asis', echo=FALSE}
models.duration <- list("Model 8" = model.d.base, "Model 9" = model.d.base2, "Model 10" = model.d.griev, "Model 11" = model.d.griev2, "Model 12" = model.d.griev3)
modelsummary(models.duration, stars = T)

@@

```{r}
models.quantity <- list("Model 8" = model.n.base, "Model 9" = model.n.base2, "Model 10" = model.n.griev, "Model 11" = model.n.griev2, "Model 12" = model.n.griev3)
modelsummary(models.quantity, stars = T)
```

Turning to models of protest duration presented in Table 3, there is no statistically significant relationship between fraud and duration in the models of economic grievance (Models 10 through 12) or in the model without any interaction term (Model 8). Only in Model 9, where the severity of fraud is interacted with the winner's vote-share, is a statistically significant relationship observed. The models suggest that the severity of election manipulation has a significant relationship with the duration of post-election protest only in extreme cases (conditional on protests having begun). However, this relationship only holds for the most extreme cases---when winning vote-shares are either very small or very large. The values of *intentional voting irregularities* used to calculate the marginal effects in Figure 4 represent one standard deviation above and below the mean for that variable. For nearly all cases, there is no expected difference in protest duration between a relatively clean and highly fraudulent election.

Three other salient observations can be taken from the marginal effects. First, the expected duration of protest remains essentially constant for a relatively fraudulent election, regardless of the margin of victory. Second, with the caveat that the result is based on a small number of observations, it appears that the expected duration of a protest movement following a narrow incumbent victory in a relatively clean election is substantially longer than would be expected in a fraudulent election. Lastly, while there is a positive expected protest duration in high-margin, high-fraud elections, the expected duration approaches zero when a high-margin election is relatively clean. Taken together, all of this supports an understanding of fraud as a noisy signal. When an incumbent soundly defeats the opposition without resorting to fraud, the incumbent is unambiguously strong---a deterrent to protest \footnote{This is not to say that the regime is unambiguously \textit{popular}, since a large body of citizens may oppose the regime but sit out the election by choice or by force.}. Likewise, when fraud is absent and the incumbent's vote-share is small, the regime is unambiguously weak. Fraud muddles the picture, however, leading to instances where organizations and citizens are uncertain of the incumbent's true strength. As a result, strong incumbents sometimes face protest movements that they can quickly demobilize. It is also possible that fraud has a protective function for incumbents in low-margin elections, by making the sources of the regime's strength uncertain, though this result has a limited base of support.

@@

We do observe support for the more structural explanations of durable protest. The coefficient for the *legislative constraint* variable, which measures the size and oppositional character of the legislative opposition, is consistently positive and  statistically significant. In Models 8 and 9 (those without a grievance interaction), the coefficient on *national party organization* is also positive and significant. The implication is that a larger, more institutionalized opposition party or coalition prolongs protest, in line with the 'test of strength' model. Similarly, the *incumbent term limit* variable is positive and consistently significant. This supports Hale's [-@hale_patronal_2014] argument that uncertainty around succession in non-democracies makes mobilization more intense as competing power centers emerge. Finally, there is some evidence in Models 9 and 11 that greater degrees of judicial constraint on the executive are associated with reduced protest duration. While not conclusive, this suggests that courts may play a role in diffusing protest, perhaps by providing parties with an alternative venue to press their claims [@Moustafa2014a;@Epperly2019c].

@@

The above results make several contributions. First, there is some evidence that worsening fraud is associated with a greater likelihood that some post-election protest will occur. In particular, there is support for the argument that fraudulent elections are associated with a greater risk of protest when the incumbent wins by a narrow margin. However, the evidence does not support the claim that fraud is more likely to catalyze protest when economic conditions are poor. This is consistent with a strategic, rather than grievance-based, model of electoral protest. In other words, the results in Table 2 support a model of electoral protest in which organizations and potential protesters use the extent of fraud as a signal of the political strength of the incumbent, an evaluation which occurs largely independent of economic grievances.

However, the results from the models of duration support the argument that election fraud is a noisy and short-lived signal of incumbent strength. In most models, the extent of election fraud has no significant association with the duration of protest. Instead, duration appears to be driven largely by institutional indicators of the balance of power between the opposition and incumbent---in particular, the size of the legislative opposition, and whether or not the chief executive is term limited. Just as in the binary models, grievances do not appear to make longer-lasting protest more likely in fraudulent elections. The results from Model 9 complicate the support for the strategic model found in Table 2, since fraud is not associated with longer protest in a low-margin election. The combination of the two results suggests that much strategic electoral protest is a gamble that opposition groups lose; anticipating a weak incumbent, they instead find one that is capable of repressing, coopting, or conciliating an emerging protest movement. Indeed, while there are 187 cases of protest out of the 647 elections in this dataset, only 15 elections resulted in either the incumbent being replaced due to protest or the election being re-run.\footnote{This information is determined using the NELDA variables NELDA37 and NELDA41. The cases of successful protest are: Ukraine 2004, Azerbaijan 2005, Georgia 2003, Kazakhstan 1994, Kyrgyzstan 2005, Bulgaria 1990, Malaysia 2004, Bangladesh 1996, Mali 1997, Peru 2000, Burkina Faso 1997, and Cote d'Ivoire 2000.}

These are pessimistic findings for proponents of democracy. While the results do find that there are circumstances in which more fraudulent elections are associated with a higher risk of *some* post-election protest occurring, this risk is limited primarily to cases in which the incumbent party wins by a narrow margin. Though protests may occur in such settings, they are no more likely to persist than in  With a large margin of victory, perhaps generated by extensive manipulation, protest appears much less likely. Moreover, the risk of protest does not appear to be heightened by poor economic conditions or intensive corruption in the public sector. As a result, circumstances that are to some extent outside of leaders' control, like a worsening macroeconomic climate, do not appear to make engaging in fraud any riskier. Of course, such conditions are likely to have an independent effect on incumbent's vote-share, but to an extent this is already 'priced in,' and incumbents may be able to effectively blunt their impact by artificially inflating their vote-share using fraud and other techniques.

More importantly, the severity of fraud appears almost wholly unrelated to the duration or number of post-election protests. Given that more widespread, frequent, or long-last protests pose a more significant challenge to non-democratic governments than isolated, rare, or brief events, these results suggest that the political cost of fraudulent elections is considerably lower than often assumed. Likewise, corruption and poor economic conditions do not appear associated with longer or more numerous protest.

Instead, the conditions that are correlated with more damaging protest are generally institutional factors, which are either more amenable to control or can be anticipated by non-democratic incumbents. For example, presidential elections appear to be associated with a greater risk of protest than legislative elections. Non-democratic leaders in presidential systems can mitigate this risk by adjusting the electoral calendar (i.e. to separate legislative and presidential elections, or by holding presidential elections early), or by ensuring that the state's coercive apparatus is fully in place prior to elections. Legislative constraints, an important factor for the duration and number of protests, can be adjusted through institutional reforms and cooptation of the opposition, over time. Finally, incumbent term limits can be managed through constitutional reforms, where possible, or through the appointment of (presumably) loyal successors. In other words, election fraud only has a minor cost in terms of protest risk, a risk which can be further blunted to the extent that elections can often be held on incumbents' terms.

None of this is to say that election fraud cannot, in theory, pose a cost to incumbents in the form of protest. This study is not a causal one; the decision by the incumbent to engage in fraud and the decision by opposition groups to protest are endogenous to one another. Instead, it shows that in practice, manipulated elections are not associated with longer or more numerous protests than clean ones. In typical cases, as a result, protest risk is likely not the constraining factor on the level of election fraud, contrary to the assumption in much of the formal theoretical literature. Rather, the constraints on manipulation are more likely to be driven by factors like resource availability [@greene_why_2007], principal-agent problems [@rundlett2016deliver], and legal risks to manipulators [@Harvey2022].

Consider also interacting fraud with rentier state status

>>>>

---
title: "Draft: It's Not So Bad---The minimal effect of election fraud on protest intensity in non-democracies"
author: "Cole J. Harvey (Oklahoma State University)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    bookdown::pdf_document2: default
linkcolor: blue
bibliography: C:/Users/colej/Documents/Research projects/bib_ch.bib
#biblio-style: "apalike"
fontsize: 12pt
link-citations: yes
toc: false
csl: C:/Users/colej/Documents/Research projects/ji_manipulation_protest/american-journal-of-political-science.csl
header-includes: \usepackage{booktabs}
abstract: "Multiparty elections and courts with a degree of independence are common features of non-democratic politics. Previous research posits a connection between these two institutions, arguing that judicial independence increases the risk of post-election protest, thus reducing electoral manipulation. This paper argues that the relationship is not straightforward, identifying four distinct (and competing) channels through which courts may affect protest risk. Using cross-national data from 1990-2012, the paper shows that independent courts only increase protest risk under limited circumstances. The results show that judicial independence increases protest risk where economic grievances are high, while reducing protest risk where grievances are low. As a result, creates judicial independence an underappreciated risk of democratic backsliding in well-off societies, while improving the odds of democratization elsewhere."
---

@@

Using cross-national data on 467 elections from 1990 to 2012, adapted from the recently-released Electoral Contention and Violence (ECAV) dataset [@Daxecker2019], this paper finds that more fraudulent elections are associated with a higher risk any post-election protest event occurring. This is in keeping with the formal-modeling literature, which predicts that overreach in the form of election fraud indicates government weakness, making opposition groups confident of winning a conflict in the streets. However, there is no statistically significant association between the level of election fraud and the number of protest events. This suggests a more complex relationship between election fraud and protest. One the one hand, the result is consistent with the prediction of some formal models that fraud is a noisy signal---in some cases, opposition groups may misperceive a strong government as weak. On the other hand, since fraud is not associated with the overall number of events in a protest wave, the risk to governments has likely been overemphasized. That is, even if protests do break out in response to large-scale fraud, the government has many tools it may deploy to curtail them. To the extent that larger protest waves are more damaging to the government than small ones, protest risk does not appear to be a substantial constraint on the severity of election fraud.

@@

Large, sustained protest waves may also contribute to elites' decisions to stand by the incumbent regime, or to side with "the people" in their own pursuit of power [@hale_patronal_2014;@svolik_politics_2012;@robertson_protest_2010].

The literature thus expects that manipulated elections increase the risk of protest by 1) revealing regime weakness, especially when the margin of victory is low, and 2) exacerbating existing grievances and upsetting citizens' sense of procedural fairness. However, as discussed below, electoral protest is relatively rare, and even in cases where it does occur, most efforts to overturn fraudulent election results through protest have failed [@kalandadze2009electoral]. The apparent failure of protest risk to deter election fraud underlines our limited understanding of non-democratic ruling parties' decision-making. Does the extent of fraud, or the severity of underlying social grievances, make more numerous (and thus more costly) protest more likely?

@@

With some exceptions, such as the study by Lankina and Skovoroda [-@Lankina2017], most research on election fraud and protest treats the latter as a binary outcome. The foregoing discussion of the literature suggests several baseline hypotheses to be tested regarding post-election protest as a binary variable. These include, first, the prediction that more severe election fraud is a signal of deeper regime weakness (i.e. unpopularity), which makes protest more likely. Second, there is the alternative view that election manipulation is most likely to spark a protest movement when margins of victory are narrow [@Simpser2011]. Finally, the grievances approach suggests that fraud will most likely lead to protest when social grievances are high [@Brancati2016;@tucker2007].

@@

However, these predictions may not apply to the size or intensity of a protest wave. Once protests begin, the extent of electoral fraud is unlikely to affect their scope. The initial emergence of protest triggers a response by the government and ruling party that can take the form of repression, concessions, or toleration [@Tilly1977;@Lipsky1968;@Hummel2019;@Rasler1996;@Bishara2015;@Frantz2014]. (**See others**.) Given that post-electoral protests are a direct challenge to the regime, toleration is unlikely in these cases. That election dates are known in advance creates opportunities for opposition mobilization; anticipating this, regimes can more easily engage in pre-emptive repression [@Truex2019;@Sullivan2016;@Bhasin2013]. Because the emergence of protest indicates that the regime's pre-emptive efforts have failed, it is not clear a priori how the regime should handle protesters it failed to deter [@Ritter2016]. In many cases, the government doubles down on repression, calling upon the security services to disperse protests, arrest leaders, and perhaps employ harsher methods including disappearance, torture, and execution [@Escriba_Folch2013;@Davenport2007]. Though such crackdowns risk provoking a backlash, they are aimed at demobilizing the opposition by increasing the costs of participation in overt dissent [@DeVogel2022]. In other cases, the government may attempt to accommodate the opposition, by making concessions on policy or institutions [@Moore2000;@Hafner-Burton2016;@Piven1979]. Alternatively, concessions may be aimed at the more moderate public, in order to keep them on the sidelines of the conflict [@Hummel2019].

@@

In this struggle between the potential protest coalition and the incumbent government, protesters and their possible allies behave as "passionate economists" [@VanZomeren2012]---they are motivated both by emotional responses to grievance, and by a more rational calculation of individual-level risk and group-level efficacy. What is the likelihood that protest action can help resolve the grievance, and does the benefit of participating outweigh the risk of personal consequences to the individual for participating?  A fraudulent election may heighten citizens' grievances and generate uncertainty around the underlying popularity of the regime, creating a moment of elevated protest risk. Crucially, though, this evaluation is dynamic and can be updated over time in response to new information and new events [@VanZomeren2012]. If protest movements have sufficient numbers and organizational resources to put the regime on the back foot, this can lead to more numerous protests as individuals' sense of group efficacy grows and their sense of risk declines [@Chenoweth_Stephan2011]. On the other hand, if the regime is well-resourced, capable, and unified, its actions to end a protest movement (including repression, concessions, and co-optation) can lead to a negative re-evaluation of the movement's odds of success. This does not, of course, eliminate citizens' sense of grievance. Instead, it leads them to abandon 'approach-oriented' forms of coping, to avoidance-oriented forms: demobilization and disengagement [@van_zomeren_put_2004;@Young2019]. For example, while ruling-party supporters in Russia disapprove of election fraud [@Reuter2021], they are more likely to withdraw from politics after exposure to electoral protest than to support the opposition [@Tertytchnaya2020].

In summary, though election fraud may be taken as a signal by opposition groups, it is a noisy one: weak and strong governments alike may engage in election manipulation. However, once a protest movement begins, the level of election manipulation is no longer a useful signal of incumbent strength. Instead, protesters and bystanders can update their estimation of the relative strength of the protest coalition and the incumbent government by observing the size of crowds, the number of protests, and the government's reaction. Since the level of manipulation will no longer serve as an important indicator of regime strength once a challenge is taken to the streets, the number of protests will instead be related to other indicators of opposition and incumbent capacity. As identified in prior research, these factors include the previous repressive history of the government [@Davenport2019], the institutionalization of opposition parties [@trejo_ballot_2014], the margin of victory in the election [@simpser2013], and whether or not the chief executive faces term limits [@hale_patronal_2014]. Generally speaking, these structural variables indicate aspects of the political opportunity structure that make protest more likely to occur [@meyer_protest_2004].

>H4: Once protests begin, the severity of election fraud will not be correlated with the number of protest events.

@@

ECAV, from which study's dependent variable is drawn, records protests and other contentious actions from six months before to three months after an election. For this study, I include protests only, excluding acts like shootings, bombings, and coups. ECAV data is in event-day format. To capture the *number of protests*, I take the sum number of events recorded in ECAV from election day to the end of the dataset's coverage period ninety days later. The result is a count variable ranging from zero to 90.\footnote{That the max value for the quantity of protests matches the number of days in the recording window is coincidental.} The large majority of elections experience no post-election protest: in 460 of the 647 elections, no protests are recorded. Figure 1 reports the distribution of protest counts across the dataset, as well as winners' vote-shares.

@@

This distribution is in line with the theoretical proposition that two data-generating processes are at work: the factors that lead to the occurrence of protest may not be the same as those that sustain or intensify it. Consequently, I model the results in two ways. The first approach uses logit models to identify the factors that make *any protest* more likely to occur---this is a binary variable that is coded as 1 if any protest occurred after the election. Since election-year observations are nested within countries here, I use multilevel models with random intercepts by country in. In the second analysis, I use negative binomial regressions to model the count of protests given that protest has begun. That is, these latter models exclude cases where no protest occurred. For ease of interpretation, these processes are modeled and depicted separately in the main text, while hurdle regression models (which use the full dataset in one stage) are presented in the appendix, with similar results.

@@

Next, all models include three measures of the degree of political openness in the country. The first of these is an indicator variable for *electoral democracy*; the reference category is electoral autocracy. The second and third are *judicial constraints* and  *legislative constraints* on the executive, V-Dem variables which captures the extent to which the judicial and legislative bodies in the country act independently in practice to limit executive power. Since urban areas can make electoral manipulation more difficult [@ziblatt2009;@larreguy2016parties] while also making protest more likely [@Wallace2013a], I also include data on *urbanization* taken from the United Nations [@DepartmentofEconomicandSocialAffairsPopulationDivision2019].

@@

```{r count-models-interact, warning = FALSE, message = FALSE, include = F}
model.d.base2 <- glm.nb(d.events.num ~ jucon.1lag + legcon.1lag + gdpgro.1lag + pubseccorrup.lag + osorg.1lag + econ.crisis.nelda.1lag + natlpartyorg.lag + v2elirreg.inv + pubseccorrup.lag +    winner.share.upd*v2elirreg.inv  + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + rentierstate + taxstate, data = protest.cases %>% filter(gdpgro.1lag < 1))
#summary(model.d.base2)
#p2 <- sjPlot::plot_model(model.d.base2, type = "pred", terms = c("winner.share.upd [all]", "v2elirreg.inv [-.36, 1.85]")) + coord_cartesian(ylim = c(0, 200))

model.n.base2 <- glm.nb(n.events.post ~ jucon.1lag + legcon.1lag + gdpgro.1lag + pubseccorrup.lag + osorg.1lag + econ.crisis.nelda.1lag + natlpartyorg.lag + v2elirreg.inv + pubseccorrup.lag +   winner.share.upd*v2elirreg.inv  + presidential + inc.termlimit.num  +  physinteg.1lag + physintlag.sq + urban.pct + rentierstate + taxstate, data = protest.cases %>% filter(gdpgro.1lag < 1))
#summary(model.n.base2)
p2 <- sjPlot::plot_model(model.n.base2, type = "pred", terms = c("winner.share.upd [all]", "v2elirreg.inv [-.36, 1.85]")) + coord_cartesian(ylim = c(0, 25))

@@

```{r, include=FALSE}
#models.quantity <- list("Model 13" = model.n.base, "Model 14" = model.n.base2, "Model 15" = model.n.griev, "Model 16" = model.n.griev2, "Model 17" = model.n.griev3)
#modelsummary(tidytest)
#modelsummary(models.quantity, stars = T)
```

@@

```{r tables-duration, results='asis', echo=FALSE, include=FALSE}
models.duration <- list("Model 8" = model.d.base, "Model 9" = model.d.base2, "Model 10" = model.d.griev, "Model 11" = model.d.griev2, "Model 12" = model.d.griev3)
modelsummary(models.duration, stars = T)

@@

```{r tables-quantityr}
models.quantity <- list("Model 8" = model.n.base, "Model 9" = model.n.base2, "Model 10" = model.n.griev, "Model 11" = model.n.griev2, "Model 12" = model.n.griev3)
modelsummary(models.quantity, stars = T)
```

Turning to models of protest quantity presented in Table 3, there is no statistically significant relationship between fraud and protest in any of the models. Figure 4, using results from Model 9, presents the marginal effects for fraud conditional on the incumbent's vote-share. As the figure shows, there is no statistically significant difference in the expected number of protest following a relatively clean election compared to a relatively fraudulent one, for any margin of incumbent victory. Taking the results from Tables 2 and 3 together, there is support for Hypothesis 1 (an unconditional increase in the risk of protest as fraud increases), as well as strong support for Hypothesis 2 (an increase in protest risk in fraudulent, low-margin elections) and Hypothesis 4 (no relationship between fraud and protest quantity). Hypothesis 3, on the connection between fraud, grievance, and protest, is not supported.

We do observe support for the more structural explanations of durable protest. The coefficient for the *legislative constraint* variable, which measures the size and oppositional character of the legislative opposition, is consistently positive and  statistically significant. In Models 8 and 9 (those without a grievance interaction), the coefficient on *national party organization* is also positive and significant. The implication is that a larger, more institutionalized opposition party or coalition prolongs protest, in line with the 'test of strength' model. Similarly, the *incumbent term limit* variable is positive and consistently significant. This supports Hale's [-@hale_patronal_2014] argument that uncertainty around succession in non-democracies makes mobilization more intense as competing power centers emerge.

@@

@@

Altogether, the findings support an understanding of fraud as a signal of incumbent strength, but a noisy and unreliable one. When an incumbent party wins by a narrow margin in a fraudulent election, the likelihood of protest initiation is substantially higher than when it wins more cleanly: roughly a 75\% chance in the former case versus 15\% in the latter as shown in Figure 2. However, since even strong incumbents commit fraud, the decision to take to the streets is often a losing gamble---anticipating a weak incumbent, they instead find one that is capable of repressing, coopting, or conciliating an emerging protest movement. Indeed, while there are 187 cases of protest out of the 647 elections in this dataset, only 15 elections resulted in either the incumbent being replaced due to protest or the election being re-run.\footnote{This information is determined using the NELDA variables NELDA37 and NELDA41. The cases of successful protest are: Ukraine 2004, Azerbaijan 2005, Georgia 2003, Kazakhstan 1994, Kyrgyzstan 2005, Bulgaria 1990, Malaysia 2004, Bangladesh 1996, Mali 1997, Peru 2000, Burkina Faso 1997, and Cote d'Ivoire 2000.} This effect can be seen in the results from the quantity models.

The above results have several implications for understanding election manipulation and regime resilience.  Most importantly, by treating protest primarily as a binary event rather, the existing literature on electoral protest has overemphasized the risk to incumbents of engaging in electoral manipulation. This results presented here show that most electoral protests quickly dwindle, and that fraud has no bearing on how numerous protest events are. Moreover, contrary to some prior research, the risk of protest does not appear to be heightened by poor economic conditions [@Brancati2014a] or intensive corruption in the public sector. As a result, circumstances that are to some extent outside of leaders' control, like a worsening macroeconomic climate, do not appear to make engaging in fraud any riskier. Instead, the conditions that are correlated with more damaging protest are generally institutional factors, which are either more amenable to control or can be anticipated by non-democratic incumbents.  Legislative constraints, an important factor for the duration and number of protests, can be adjusted through institutional reforms and cooptation of the opposition, over time. Finally, incumbent term limits can be managed through constitutional reforms, where possible, or through the appointment of (presumably) loyal successors. In other words, election fraud only has a minor cost in terms of protest risk, a risk which can be further blunted to the extent that elections can often be held on incumbents' terms. These are pessimistic findings for proponents of democracy.

That electoral protest is less risky than is often assumed, in turn, implies that governments and ruling parties have a freer hand to engage in electoral manipulation than many models posit. Since the severity of fraud has no apparent bearing on whether or not a major protest wave will emerge as a serious threat to the regime, as shown in Figure 3, government have an incentive to manipulate as much as possible to secure a large winning vote-share. This perverse incentive undermines the case that democracy, and fair elections generally, are a self-enforcing equilibrium backed up by the risk of citizen protest [@fearon_self-enforcing_2011].

None of this is to say that election fraud cannot, in theory, pose a cost to incumbents in the form of protest. This study is not a causal one; the decision by the incumbent to engage in fraud and the decision by opposition groups to protest are endogenous to one another. Instead, it shows that in practice, manipulated elections are not associated with more numerous protests than clean ones. In typical cases, as a result, protest risk is likely not the constraining factor on the level of election fraud, contrary to the assumption in much of the formal theoretical literature. Rather, the constraints on manipulation are more likely to be driven by factors like resource availability [@greene_why_2007], principal-agent problems [@rundlett2016deliver], and legal risks to manipulators [@Harvey2022].

# Conclusion

Protest risk is often considered to be a major risk to incumbent governments that work to undermine election integrity. This risk has a central role in several prominent models of electoral manipulation, serving to deter manipulation and to incentivize the toleration of election monitors. However, empirical research on the subject has largely treated protest as a binary variable, which obscures the degree to which post-election protest imposes a real risk on incumbents in non-democracies. One protest may be easily contained, while dozens may pose a more significant risk to the regime.

This paper addresses this concern by testing the relationship between election fraud, the risk that any protest occurs, and the number of protest events that occur after an election. Drawing on data from the V-Dem, NELDA, and ECAV datasets, it uses cross-national data on 647 elections to show that---while there is an elevated risk of some protest occurring in a fraudulent election with narrow margins of victory---election fraud has no significant relationship with the number of protests that occur in a protest wave. These results imply that protest risk should not be considered a central deterrent to election fraud. Instead, most post-election protests are strategic gambles gone wrong; the initiation of protest more often reveals the incumbent to be strong enough to manage protest through repression, cooptation, or sideways concessions. These findings indicate that, rather than being deterred by protest risk, governments have an incentive to manipulate as much as possible in order to achieve a large official vote-share. The evidence suggests that democratic breakthoughs in non-democratic elections are rare and unlikely to be driven by the severity of manipulation---a gloomy picture for supporters of democratization---and that researchers should consider other mechanisms beyond mass protest in their models of electoral manipulation.
